---
title: "Assignment 3"
author: "ALZ Alliance"
format:
  html:
    toc: true
execute: 
  warning: false
---

```{r}
#| echo: false

knitr::opts_chunk$set(
  eval = TRUE,
  message = TRUE,
  warning = TRUE,
  error = TRUE
)
```

```{r}
# Loading the required library
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
```

## Data processing

```{r}
# loading the data 
startup_finance <- read_csv(here::here("data","Finance_and_Insurance.csv")) |> clean_names()
startup_food <- read_csv(here::here("data","Food_and_services.csv")) |> clean_names()
startup_health <- read_csv(here::here("data","Health_Care.csv")) |> clean_names()
startup_it <- read_csv(here::here("data","Information_Sector.csv")) |> clean_names()
startup_manufacture <- read_csv(here::here("data","Manufactures.csv")) |> clean_names()
startup_retail <- read_csv(here::here("data","Retail_Trade.csv")) |>  clean_names()
```

#### Combining dataset into one

```{r}
# combining the dataset into one

startup_combined <- bind_rows(
  startup_finance,
  startup_food,
  startup_health,
  startup_it,
  startup_manufacture,
  startup_retail
)
```

#### Data cleaning

-   Convert years of operation into `start_year` and `end_year`
-   Mutate a col for length of business
-   Convert `how_much_they_raised` into numeric

#### Understanding Dataset

```{r adding table for data description}

variable_description <- tibble::tibble(
  Variable = c(
    "name", "sector", "years_of_operation", "what_they_did",
    "how_much_they_raised", "why_they_failed", "takeaway", "giants",
    "no_budget", "competition", "poor_market_fit", "acquisition_stagnation",
    "platform_dependency", "monetization_failure", "niche_limits", "execution_flaws",
    "trend_shifts", "toxicity_trust_issues", "regulatory_pressure", "overhype"
  ),
  Description = c(
    "Name of the startup",
    "Business sector or industry",
    "Number of years the startup operated",
    "Description of what the startup did",
    "Total funds raised before failure (if available)",
    "Primary reason for startup failure",
    "Key takeaway or lesson from the failure",
    "Whether the startup competed with major players",
    "Failed due to budget constraints",
    "Faced tough competition",
    "Did not fit well with market demand",
    "Growth plateaued or stagnated",
    "Depended heavily on external platforms",
    "Could not effectively monetize",
    "Had limited market potential",
    "Poor execution or management issues",
    "Impact of changing industry trends",
    "Trust or toxicity issues in the team",
    "Faced legal or regulatory hurdles",
    "Failure due to excessive hype or unrealistic expectations"
  )
)

# Display as a table
kbl(variable_description, caption = "Variable Descriptions for Startup Combined Dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

## Executive Summary (Linh)

## Introduction (Linh)

## Methodology (Ayush)

## Results (Zul + Linh)

### Why startups fail (Linh)

```{r}

# Check unique names
```

### Temporal analysis of start-up failures (Zul)

## Conclusion (Zul + Linh)

### Why startups fail (Linh)

### Temporal analysis of start-up failures (Zul)