---
title: "Assignment 3"
author: "ALZ Alliance"
format:
    html:
        toc: true
editor: visual
execute: 
  warning: false
---

```{r}
#| echo: false

knitr::opts_chunk$set(
  eval = TRUE,
  message = TRUE,
  warning = TRUE,
  error = TRUE
)
```


```{r}
# Loading the required library
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
```

## Data processing

```{r}
# loading the data 
startup_finance <- read_csv(here::here("data","Finance_and_Insurance.csv")) |> clean_names()
startup_food <- read_csv(here::here("data","Food_and_services.csv")) |> clean_names()
startup_health <- read_csv(here::here("data","Health_Care.csv")) |> clean_names()
startup_it <- read_csv(here::here("data","Information_Sector.csv")) |> clean_names()
startup_manufacture <- read_csv(here::here("data","Manufactures.csv")) |> clean_names()
startup_retail <- read_csv(here::here("data","Retail_Trade.csv")) |>  clean_names()
```

#### Combining dataset into one

```{r}
# combining the dataset into one

startup_combined <- bind_rows(
  startup_finance,
  startup_food,
  startup_health,
  startup_it,
  startup_manufacture,
  startup_retail
)
```

#### Data cleaning

- Convert years of operation into `start_year` and `end_year`
- Mutate a col for length of business
- Convert `how_much_they_raised` into numeric

##### Remove duplicated rows

```{r find duplicated names and remove them}

# Find companies with multiple observations
duplicated_names <- startup_combined |> 
  count(name) |> 
  filter(n > 1) |> 
  pull(name)

# Take a look to pick out which rows to remove
startup_combined |> 
  mutate(id = row_number()) |> 
  filter(name %in% duplicated_names) |> 
  arrange(name)

# Handpicked the rows to remove
removed_rows <- c(264, 89, 175, 306, 66, 259)
startup_combined <- startup_combined |> 
  mutate(id = row_number()) |> 
  filter(!id %in% removed_rows)
```

#### Understanding Dataset

```{r}
# adding table for data description

variable_description <- tibble::tibble(
  Variable = c(
    "name", "sector", "years_of_operation", "what_they_did",
    "how_much_they_raised", "why_they_failed", "takeaway", "giants",
    "no_budget", "competition", "poor_market_fit", "acquisition_stagnation",
    "platform_dependency", "monetization_failure", "niche_limits", "execution_flaws",
    "trend_shifts", "toxicity_trust_issues", "regulatory_pressure", "overhype"
  ),
  Description = c(
    "Name of the startup",
    "Business sector or industry",
    "Number of years the startup operated",
    "Description of what the startup did",
    "Total funds raised before failure (if available)",
    "Primary reason for startup failure",
    "Key takeaway or lesson from the failure",
    "Whether the startup competed with major players",
    "Failed due to budget constraints",
    "Faced tough competition",
    "Did not fit well with market demand",
    "Growth plateaued or stagnated",
    "Depended heavily on external platforms",
    "Could not effectively monetize",
    "Had limited market potential",
    "Poor execution or management issues",
    "Impact of changing industry trends",
    "Trust or toxicity issues in the team",
    "Faced legal or regulatory hurdles",
    "Failure due to excessive hype or unrealistic expectations"
  )
)

# Display as a table
kbl(variable_description, caption = "Variable Descriptions for Startup Combined Dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

## Executive Summary (Linh)

## Introduction (Linh)

Startups have long been engines of innovation, yet their failure rate remains notoriously high — more than two-thirds of them are unable to generate positive returns (HBR). In recent years, macroeconomic challenges and a venture funding freeze since 2022 have made startup survival even more difficult (source). This environment raises an urgent question: **Why and when do startups fail, and how do factors like industry and funding levels influence these outcomes?** To answer this, we analyzed post-mortems from 483 failed startups compiled by CB Insights, where founders shared the reasons behind their collapse (source). These candid reflections offer rare, unfiltered insights into the internal and external pressures startups face. 

Our analysis focuses on two key dimensions: (1) the reasons for failure (broken down by sector and funding levels), and (2) a temporal analysis to identify shifts in causes over time. Understanding these patterns can help startup founders, investors, and policymakers better anticipate pitfalls and design more resilient strategies. Ultimately, this research aims to uncover not only on what goes wrong — but also when and for whom — to inform smarter decision-making in the startup ecosystem.

## Methodology (Ayush)

## Results (Zul + Linh)

### Why startups fail (Linh)

### Temporal analysis of start-up failures (Zul)

## Conclusion (Zul + Linh)

### Why startups fail (Linh)

### Temporal analysis of start-up failures (Zul)
