---
title: "Assignment 3"
author: "ALZ Alliance"
format:
  html:
    toc: true
execute: 
  warning: false
---

```{r}
#| echo: false

knitr::opts_chunk$set(
  eval = TRUE,
  message = TRUE,
  warning = TRUE,
  error = TRUE
)
```

```{r}
# Loading the required library
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(ggplot2)
```

## Data processing

```{r}
# loading the data 
startup_finance <- read_csv(here::here("data","Finance_and_Insurance.csv")) |> clean_names()
startup_food <- read_csv(here::here("data","Food_and_services.csv")) |> clean_names()
startup_health <- read_csv(here::here("data","Health_Care.csv")) |> clean_names()
startup_it <- read_csv(here::here("data","Information_Sector.csv")) |> clean_names()
startup_manufacture <- read_csv(here::here("data","Manufactures.csv")) |> clean_names()
startup_retail <- read_csv(here::here("data","Retail_Trade.csv")) |>  clean_names()
```

#### Combining dataset into one

```{r}
# combining the dataset into one

startup_combined <- bind_rows(
  startup_finance,
  startup_food,
  startup_health,
  startup_it,
  startup_manufacture,
  startup_retail
)
```

#### Data cleaning

-   Convert years of operation into `start_year` and `end_year`
-   Mutate a col for length of business
-   Convert `how_much_they_raised` into numeric

##### Remove duplicated rows

```{r find duplicated names and remove them}

# Find companies with multiple observations
duplicated_names <- startup_combined |> 
  count(name) |> 
  filter(n > 1) |> 
  pull(name)

# Take a look to pick out which rows to remove
startup_combined |> 
  mutate(id = row_number()) |> 
  filter(name %in% duplicated_names) |> 
  arrange(name)

# Handpicked the rows to remove
removed_rows <- c(264, 89, 175, 306, 66, 259)
startup_combined <- startup_combined |> 
  mutate(id = row_number()) |> 
  filter(!id %in% removed_rows)
```

#### Understanding Dataset

```{r adding table for data description}

variable_description <- tibble::tibble(
  Variable = c(
    "name", "sector", "years_of_operation", "what_they_did",
    "how_much_they_raised", "why_they_failed", "takeaway", "giants",
    "no_budget", "competition", "poor_market_fit", "acquisition_stagnation",
    "platform_dependency", "monetization_failure", "niche_limits", "execution_flaws",
    "trend_shifts", "toxicity_trust_issues", "regulatory_pressure", "overhype"
  ),
  Description = c(
    "Name of the startup",
    "Business sector or industry",
    "Number of years the startup operated",
    "Description of what the startup did",
    "Total funds raised before failure (if available)",
    "Primary reason for startup failure",
    "Key takeaway or lesson from the failure",
    "Whether the startup competed with major players",
    "Failed due to budget constraints",
    "Faced tough competition",
    "Did not fit well with market demand",
    "Growth plateaued or stagnated",
    "Depended heavily on external platforms",
    "Could not effectively monetize",
    "Had limited market potential",
    "Poor execution or management issues",
    "Impact of changing industry trends",
    "Trust or toxicity issues in the team",
    "Faced legal or regulatory hurdles",
    "Failure due to excessive hype or unrealistic expectations"
  )
)

# Display as a table
kbl(variable_description, caption = "Variable Descriptions for Startup Combined Dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

## Executive Summary (Linh)

## Introduction (Linh)

## Methodology (Ayush)

## Results (Zul + Linh)

### Why startups fail (Linh)

#### Overall reason
```{r overall failure reasons}
col_pivot <- c("giants", "no_budget", "competition", "poor_market_fit", "acquisition_stagnation",
"platform_dependency", "monetization_failure", "niche_limits", "execution_flaws",
"trend_shifts", "toxicity_trust_issues", "regulatory_pressure", "overhype", "high_operational_costs")

failure_counts <- startup_combined |> 
  select(col_pivot) |> 
  mutate_all(~replace_na(., 0)) |> 
  pivot_longer(cols = col_pivot, names_to = "reason", values_to = "present") |> 
  group_by(reason) |> 
  summarise(count = sum(present)) |> 
  ungroup() |> 
  mutate(percent = (count/nrow(startup_combined))*100)
```

```{r plot overall failure reason}
#| label: fig-overall-reason
#| fig-cap: "The majority of startups failed due to intense competition, especially with giant players in the market."

col_pivot_labels <- c(
  giants = "Giants",
  no_budget = "No budget",
  competition = "Competition",
  poor_market_fit = "Poor market fit",
  acquisition_stagnation = "Acquisition stagnation",
  platform_dependency = "Platform dependency",
  monetization_failure = "Monetization failure",
  niche_limits = "Niche limits",
  execution_flaws = "Execution flaws",
  trend_shifts = "Trend shifts",
  toxicity_trust_issues = "Toxicity & trust issues",
  regulatory_pressure = "Regulatory pressure",
  overhype = "Overhype",
  high_operational_costs = "High operational costs"
)

failure_counts |> 
  ggplot(aes(x = reorder(reason, count), y = count)) +
  geom_bar(stat = "identity", fill = "#2166ac") +
  geom_text(aes(label = paste0(round(percent, 1), "%")), 
            hjust = -0.1,
            color = "black", size = 3) +
  coord_flip() +
  labs(title = "Count of Startups by Failure Reason",
    x = "Failure Reason",
    y = "Number of Startups",
    fill = "Sector") +
  scale_x_discrete(labels = col_pivot_labels) +
  theme_minimal() +
  theme(panel.grid = element_blank())

```

#### Reasons by sector

```{r percentage of failures by sector}
#| label: fig-sector-percent
#| fig-cap: "Besides universal reasons of Competition and Giants, each sector carried their own set of reasons for startup failures."

sector_counts <- startup_combined |> 
  group_by(sector) |> 
  summarise(sector_counts = n()) |> 
  ungroup() 

sector_percent <- failure_counts |> 
  group_by(reason, sector) |> 
  summarise(company_fail_counts = sum(count, na.rm = TRUE)) |> 
  ungroup() |> 
  left_join(sector_counts, by = "sector") |> 
  mutate(percent_fail = (company_fail_counts/sector_counts)*100)

sector_percent |> 
  ggplot(aes(x = reorder(reason, company_fail_counts), y = sector, fill = percent_fail)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(percent_fail, 1)), size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "#f0f0f0", high = "#2166ac", name = "% Failed") +
  theme_minimal() +
  labs(title = "Startup Failure Reasons by Sector",
       x = "Failure Reason",
       y = "Sector") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

```

#### Funding

```{r funding by sector}
#| eval: false

startup_combined <- startup_combined |>
  mutate(funding_millions = str_extract(how_much_they_raised, "\\d+\\.?\\d*") |> as.numeric())

startup_combined |>
  ggplot(aes(x = sector, y = funding_millions)) +
  geom_violin(fill = "#2166ac") +
  labs(
    title = "Funding Raised by Sector",
    x = "Sector",
    y = "Funding Raised (Millions)"
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)  )

```

```{r funding by reason}
#| label: fig-funding-reason
#| fig-cap: "Low-funded startups struggled with scalability of niche markets, product-market fit and financial capacity, while those with medium and high fundings faced challenges in monetizations, business executions and post-acquisition operations. "

startup_combined <- startup_combined |>
  mutate(funding_millions = str_extract(how_much_they_raised, "\\d+\\.?\\d*") |> as.numeric())

threshold <- quantile(startup_combined$funding_millions, probs = c(0.5, 0.75), na.rm = TRUE)

startup_combined <- startup_combined |> 
  mutate(funding_group = cut(funding_millions, breaks = c(-Inf, 20, 30, 70, Inf),   
        labels = c("seed","A", "B", "C"))) 
        
funding_group_counts <- startup_combined |> 
  group_by(funding_group) |> 
  summarise(funding_group_counts = n())

funding_group_percent <- startup_combined |> 
  select(col_pivot, funding_group) |>
  pivot_longer(cols = col_pivot, names_to = "reason", values_to = "present") |> 
  group_by(reason, funding_group) |> 
  summarise(count = sum(present, na.rm = TRUE)) |> 
  ungroup() |> 
  left_join(funding_group_counts, by = "funding_group") |> 
  mutate(percent_fail = (count/funding_group_counts)*100) |> 
  filter(!is.na(funding_group))
  
funding_group_percent |>
  ggplot(aes(x = reorder(reason, percent_fail), y = percent_fail)) +
  geom_col(fill = "#2166ac") +
  geom_text(aes(label = paste0(round(percent_fail, 1), "%")), 
            hjust = -0.1, color = "black", size = 3) +
  coord_flip() +
  facet_wrap(~funding_group, ncol = 4) +
  scale_x_discrete(labels = col_pivot_labels) + 
  labs(title = "Startup Failure Reasons by Funding Group",
    x = "Failure Reason",
    y = "Failure Percentage") +
  theme_minimal(base_family = "sans") +
  theme(panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(face = "plain")) +
  expand_limits(y = max(funding_group_percent$percent_fail) * 1.1)

```

### Temporal analysis of start-up failures (Zul)

## Conclusion (Zul + Linh)

### Why startups fail (Linh)

### Temporal analysis of start-up failures (Zul)
